---
title: "GNRS R package"
author: "Brian Maitner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GNRS R package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Geographic Name Resolution Service

The package `GNRS` is designed to interact with the Geographic Name Resolution Service of the Botanical Information and Ecology Network.  


#Installing GNRS
```{r, eval=FALSE}

library(devtools)
install_github("EnquistLab/RGNRS/GNRS")


```

#The easiest case: one political division

```{r}
library(GNRS)
GNRS_super_simple(country =  "United States", 
                 state_province = "Arizona",
                 county_parish = "Pima County")


```

#Multiple political divisions

```{r}

#First, we'll load the test data that are included with this package

gnrs_testfile <- read.csv(system.file("extdata", "gnrs_testfile.csv",
                                      package = "GNRS", mustWork = TRUE),
                          stringsAsFactors = FALSE)


head(gnrs_testfile, n = 10)
```

As you can see, the sample data include spelling variants (USA vs U.S.A.) and non-standard characters that may cause problems.  The GNRS will standardize these spelling variants and non-standard characters.


```{r}

gnrs_results <- GNRS(gnrs_testfile)

#The standardized names are found in these columns:
head(gnrs_results[c("country","state_province","county_parish")], n = 10)

```


The GNRS function expects 4 columns as input, but all are optional. If you ever forget, you can use the function GNRS_template as a quick look-up, or as a template to populate

```{r}


head(GNRS_template())


```

