---
title: "Resolving political division names offline"
author: "Brian Maitner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Resolving political division names offline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Status: beta

The offline engine is a new port of the resolution cascade the Geographic Name
Resolution Service runs.  It resolves against the service's own reference
tables, so identifiers, standard names and codes are the service's.  Measured
against the web service on 2026-09-03:

| set | rows | country | state/province | county/parish |
|---|---|---|---|---|
| `gnrs_testfile` | 21 | 21 of 21 | 21 of 21 | 21 of 21 |
| hand-picked edge cases (codes, alternate names, misspellings, the cross-level routes) | 179 | 177 of 179 | 178 of 179 | 178 of 179 |
| names drawn from the reference with perturbations | 1,088 | 99.4% | 96.4% | 99.4% |

The figures are agreement on the identifier of the resolved division.  Nearly
every disagreement is one of three things, discussed below: the local
alternate names are newer than the service's copy, and match where the service
does not; the service picked one of several equally good matches arbitrarily
and this package picked the other; or the service answered from its cache.
The last row is the harshest test, because a third of it submits states by an
alternate name sampled from the current GeoNames file, which the service's
2020 copy often lacks; on the rest of that set agreement is above 99% at every
level.

Beyond these sets the engine has not been tested broadly, nor independently
reviewed.  Treat the results as provisional: for work where the answer matters,
resolve a sample through `GNRS()` as well and compare, and please report
anything that disagrees.

## Why resolve names offline

`GNRS()` sends your names to the web service and returns the server's answer.
That is the right choice most of the time, but it needs an internet
connection, it needs the service to be up, and it accepts at most 5,000 rows
per request.

`GNRS_local()` runs the same resolution cascade on your own machine, against a
copy of the reference data that you download once.  It works with no network,
it does not depend on the service being available, and there is no row limit.

The two are separate functions rather than one function with a switch.  That
is deliberate: they do not always give the same answer, and it should always be
obvious from your code which one produced a result.

## One-off setup

```{r}
library(GNRS)

GNRS_local_build()
```

This fetches two things.  The first is the web service's own reference tables,
through its API: every country, state/province and county/parish it knows,
with the identifiers, standard names, codes and GADM identifiers it returns.
About 25 MB is transferred and a few megabytes kept.  The second is the
alternate-names file from GeoNames, about 195 MB, from which the service takes
the names in other languages, abbreviations and historical names its
alternate-name matching uses.  The API does not serve these, so they come from
the publisher; only the rows for the reference divisions are kept, and the
archive is deleted once they are extracted.  The whole build took about four
minutes and left 8.5 MB on disk.

To see what you have:

```{r}
GNRS_local_status()
```

```
    source                          full_name built
1     gnrs GNRS reference political divisions  TRUE
2 geonames           GeoNames alternate names  TRUE
                                  version downloaded size_mb download_mb
1 database 2.2.5 (2024-04-02), code 1.7.4 2026-09-03     5.4          25
2                              2026-09-03 2026-09-03     2.7         195
```

Keep those version strings.  They are what makes a local result citable and
reproducible: the first is the web service's database and code version at the
time of the build, the second the date of the GeoNames file.
`GNRS_local_citations()` writes them into citations for you, optionally as
BibTeX.

Asking `GNRS_local()` for data you have not built does not quietly start a
download.  It tells you what is missing and the call that would fix it; in an
interactive session you can pass `build_missing = TRUE` and answer the prompt,
and in a script the same argument builds without asking.
`GNRS_local_remove()` deletes the cache if you want the space back.

## Resolving names

```{r}
results <- GNRS_local(gnrs_testfile)

results[1:4, c("country", "state_province", "county_parish", "overall_score", "match_status")]
```

```
        country     state_province    county_parish overall_score  match_status
1        Russia Lipetskaya Oblast' Dobrovskiy Rayon             1    full match
2        Mexico             Sonora                              1 partial match
3     Guatemala             Izabal                              1    full match
4 United States            Arizona             Pima             1    full match
```

The input is the same four-column data.frame `GNRS()` takes (see
`GNRS_template()`), and the output has the same columns in the same order, so
code that reads the result of one will read the other.  Two details differ:
the scores are numeric rather than the character strings the service returns,
and an unmatched or unsubmitted value is an empty string, as the service
returns it.

The first call in a session loads the reference data, which takes a few
seconds; after that it is held in memory.  On one machine, 1,088 rows resolved
in about 20 seconds once loaded, and the web service took a few seconds for
the same rows, so for a short list of names the service is quicker.  The local
version pays off when the service is down, when there is no connection, and
for long lists, which it takes in one go.

## How local results differ from the web service

They usually agree, but not always, and the differences are worth
understanding before you rely on either.

### The alternate names are newer

The service's alternate names are from a GeoNames dump of April 2020.
`GNRS_local_build()` downloads the current file, so a name added to GeoNames
since then matches locally and not on the service.  Every disagreement of this
kind seen in validation was an improvement: "Eswatini" now resolves to the
country the service still lists as Swaziland, "DRC" to the Democratic Republic
of the Congo, "Bengaluru Urban" to the district the service only finds under
its older spelling.  It also means a name can match by a different route:
where the service finds "Praha" through a wildcard search, the local version
has it as an exact alternate name.  The resolved division is the same; the
`match_method_*` column differs.

Expect this effect to grow, and report the GeoNames file date from
`GNRS_local_status()` with your results.

### Ties are broken by identifier

The upstream SQL updates a row from a join, and when several reference rows
match equally well PostgreSQL uses whichever it visits first.  Two counties of
Punjab are both called "Lahore" in short form; the service returned one, this
package the other.  "Hbeei Sheng" is equally similar to Hebei and Hubei; the
service happened to pick Hebei.  This package always takes the lowest
identifier, so its answer is at least reproducible.  Three of the upstream
fuzzy steps also fail to select the best-scoring candidate at all, updating
with any candidate above the threshold; this package takes the best, which is
what the other steps do and what the scripts evidently intend.

### Ambiguous wildcard matches are refused

One of the service's steps strips words such as "Province of" from the
submitted name and looks for the remainder inside the alternate names.  It is
written to accept the result only when it is unique, but its grouping makes
that test ineffective, so the service takes whatever candidate comes first.
Submitted "Province of Parana" under Brazil, the remainder "Parana" contains
"Para" and is contained in "Paraná", and the service answered Pará, the wrong
state.  This package requires a unique candidate, as the step intends, and
leaves the state unmatched instead.

### No result cache

The service caches earlier answers by the trimmed submitted names, and a cache
hit returns the earlier verbatim values and methods.  That is why " Mexico "
can come back from the service as "Mexico" with an exact match while
"  Mexico  " comes back untrimmed with a fuzzy one.  This package resolves
every submission afresh.

### Other small points

A percent sign in a submitted name is treated as text rather than as the
wildcard the SQL would take it for.  HASC codes are matched in full ("US.AZ")
but not in their short form, because the deployed service does not match the
short form either: its legacy source for those codes never populated the
database, as the service's own answers show.

## Building without the alternate names

```{r}
GNRS_local_build("gnrs")
results <- GNRS_local(my_names, alternate_names = FALSE)
```

This is quick and small, but confines matching to the standard names, ASCII
forms, short forms and codes, so a country submitted in another language or a
state by its local name will not resolve.  The default builds both.

## Citing a local result

```{r}
GNRS_local_citations()
```

This lists the GNRS publication, this package, and the two data components at
the versions actually built.  The reference tables are the web service's, built
from GADM, GeoNames and Natural Earth; the alternate names are GeoNames', under
CC BY 4.0.  Please cite them: the service could not exist without them.
